Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-04-19 paper_jasa.tex
final versions
cleaning latex codes

%\begin{multicols}{2}
%\twocolumn

@@

%\checkmark

@@

%Statistical graphics research and development is \red{agreeably established} now, and has been cemented by s
Several new developments in graphics research have been achieved in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. The research by \cite{heer:2010} used subjects recruited from Amazon's Mechanical Turk \citep{turk} for their studies. This body of work provides a contemporary framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

%, analysis requires plots of data. For exploratory data analysis, statistical graphics play an invaluable role in model checking and diagnostics. Even though we have established mathematical procedures to obtain various statistics, we need to support the results by also producing the relevant plots.

%The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS ease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}.   \blue {Graham Wills has implemented the grammar in SPSS, but I'm not sure how to cite this.}

%\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between conventional statistical inference procedures and exploratory data analysis.

%In this paper we compare the lineup protocol to conventional statistical inference, specifically applied to the regression framework. We also present results of a human-subject study assessing the performance of individuals on lineup plots \citep{buja:2009} for testing significance of regression parameters.

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual~$i$ picks out the actual data plot. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot.

If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p_i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition (see \citet{butler93} for details). Ways to estimate $p_i$ will be  discussed in the following sections.
%one of the main concerns for the remainder of the paper. One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

%\green{WHY??? And how does this relate to the logistic model below?} 
%This leads to the definition of the power of the visual test, as follows:

%The lineup plot can be evaluated by one or more individuals. When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

 %Notice that when $N=1$, this $p$-value is $\frac1m$.

%\subsection{Power}

%Next, we will develop power theoretically and then relate it to the empirical results.
%We will first theoretically define and  derive the power of  a visual test in the setting of a parametric hypothesis test, i.e. we assume a test  of $H_0: \theta = 0$ vs $H_1: \theta \neq 0$ for some real valued parameter vector $\theta$. In a next step we will then relate the theoretical results to our empirical findings.

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test, $V_{\theta}$, is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0 \mid \theta) 
%      = P(X \ge x_{\alpha}) =
%        \begin{cases} 
%               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true}.\\
%              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
%        \end{cases}
    \end{equation*}
\end{dfn}

@@

%\begin{proof}
%%We will now make use of properties of the data going into the lineup and assume that the properties are reflected in the lineup display:
%
% By definition  
%
%$$p_D=Pr\left(|t| \ge t_{obs} \mid H_0\right)=1-F_{|t|}(t_{obs}) \ \ \Rightarrow \ \  |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the conventional test statistic.
%%, with non-centrality parameter $\delta$. 
%\noindent Then the distribution function of the $p$-value, $p_D$, under $H_0$, is uniform, since:  
%\begin{eqnarray}\label{dist_p}
%F_{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
%  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
%  &=&%\left\{ \begin{array}{ll}
%          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
% %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%\end{eqnarray}
%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density of $p_D$ $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%
%
%
%Let $p_{0,i}$, $i=1, ..., m-1$ denote the  $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis,  the $p$-values are independent and  follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$. The minimum $p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}$ then follows a Beta distribution with shape parameters 1 and $m-1$, and corresponding distribution function 
%\[
%F_{p_0} (x) =  1- (1- x) ^{m-1}  \text{ for } x \in [0,1].
%\]
%
%\noindent Thus
%\begin{eqnarray*}
%P(p_D < p_{0}) &=& 1 - P(p_{0} \le p_D) = 1- \int_0^1  P(p_{0} \le p_D \mid p_D=t) f_{p_D}(t) dt   \\
%&=& 1 - \int_0^1 F_{p_{0}}(t) f_{p_D}(t) dt = 1 - \int_0^1 f_{p_D}(t) dt + \int_0^1 (1-t)^{m-1} f_{p_D}(t) dt  \\
%&=&  E\left[ (1 - p_D)^{m-1}\right].
%\end{eqnarray*}
%
%
%%\noindent This proves the statement above. 
%%We further see:
%%\[
%%E\left[ (1 - p_D)^{m-1}\right] =  \sum_{k=0}^{m-1} {m-1 \choose k} (-1)^k E[p_D^k] = 1-(m-1)E[p_D] + O(E[p_D^2]).
%%\]
%
%
%% If the density of $p$-value is very right skewed, the expectation term would be large. The distribution function would be highly right skewed when there would be strong signal in the data plot.Thats where the distribution function of $p$-value under alternative comes in. Should we add that in Equation \ref{dist_p}? )},
%%conversely, as $m$, the number of choices given in the lineup, goes up the probability to pick the data plot  goes down. 
%%}
%%\blue{This shows that the lineup size $m$ has an effect on the power of the visual test: it is the $m-1$th moment of the distribution of $1-p_D$. 
%
%%
%%
%%
%%Let us think of a lineup as a head-to-head comparison of the test statistic and $m-1$ null plots. 
%%%Let the data plot have a $p$-value of $p_D$, and the null plots $p$ values of $p_{0, i}$ with $1 \le i \le m-1$.
%%We know that for each comparison the probability that the data, on which the plot is based, has the smaller $p$ value is 
%%\begin{eqnarray*}
%%P(p_D < p_{0,i}) &=& 1 - P(p_{i,0} \le p_D) = 1- \int_0^1  P(p_{i,0} \le p_D \mid p_D=t) f_{p_D}(t) dt =  \\
%%&=& 1 - \int_0^1 F_{p_{0,i}}(t) f_{p_D}(t) dt = 1 - \int_0^1 t f_{p_D}(t) dt = 1 - E[p_D],
%%\end{eqnarray*}
%%which, in particular, is independent of $p_{0,i}$ for all $i$.
%%
%%Let us make the assumption that an observer is able to identify  the chart corresponding to the data with the smallest $p$-value. Further we will assume that all observers  have the same ability in identifying the data plot.
%%
%%With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.  
%%
%%Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an observer will pick the data plot in a given lineup is 
%%\[
%%P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
%%\]
%\end{proof}

%\red{pull the results for $P(p_D < p_{i,0})$ and $P(p_D < p_0)$ to the front, including the Figure on $1/m$ versus power. These results are independent of the regression setting - they don't make any assumptions in terms of regression.}

The above lemma allows two immediate conclusions for the use of lineups.
The probability that the observer correctly identifies the data plot is closely connected to the size of the lineup $m$, since the right hand side of the above equation decreases for larger $m$, the probability of correctly identifying the actual data plot decreases with $m$. %This may seem counterintuitive, but it makes sense. 
%With fewer null plots the actual data plot will stand out, when the alternative hypothesis is true. 
Further we see, that the rate of this decrease depends strongly on the distribution of $p_D$ -- if the density of $p_D$ is very right skewed, the  expectation term on right hand side will be large and less affected by an increase in $m$. This can also be seen in Figure \ref{fig:pval_power}, which illustrates lemma \ref{lemma}.   Figure \ref{fig:pval_power} shows the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. It can also be seen that for a $p$-value, $p_D$, of about 0.15 for the data plot, the signal in the plot is so weak, that it cannot be distinguished from null plots in a lineup of size $m=20$. % This pattern is confirmed by our findings described in Section~\ref{sec:simulation}.
%also seen in our experiments as shown in Figure \ref{fig:pval_pcorrect}.

%Even under $H_1$ we expect some observers to pick a null plot with a probability that depends on the strength of the signal in our data plot. 
%Reversely, we will now make use of the number of observers who do not pick the data plot in a lineup to infer the strength of the signal in the test statistic.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. 
   %Comparison of probabilities to pick the actual data plot for different size lineups, $m$, and different values for data plot strength, $p_D$. The dashed  line, and grey 95\% pointwise confidence band, shows the probability of picking the data plot, by chance, if $H_0$ is true. }
}
   \label{fig:pval_power}
\end{figure}

@@

%A lineup including the test statistic  for a binary $X_k$ is shown in Figure \ref{fig:test_category}. The 19 null plots are generated  by simulating residuals from $N(0,\hat{\sigma}^2)$. The test statistic, the plot containing the actual data, is randomly placed among these null plots. If the test statistic is identifiable the null hypothesis is rejected. % with a $p$-value of at most 0.05.

%\subsection{Expected power}

% Di: This is specific for the regression model
%Consider the hypothesis test $H_0: \beta_k=0$ against $H_1: \beta_k \ne 0$ in model \ref{multi}.
%%Now consider estimating the power of the visual test. 
%%Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$. For the regression slope parameter $\beta$ suppose the associated test statistic in conventional test be $t_{obs}$ with $p$-value $p_D$. By definition we have 
%%
%%$$p_D=Pr(|t| \ge t_{obs}| H_0)=1-F_{|t|}(t_{obs}) \Rightarrow |t_{obs}|=F_{|t|}^{-1}(1-p_D)$$
%%
%%\red{HH: the following paragraph should probably move up, but I'm not sure how to change it appropriately. Mahbub, could you take care of that?}
%%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$ distribution with non-centrality parameter $\delta$. Thus the distribution function of $p-value$ be  
%%\begin{eqnarray}\label{dist_p}
%%F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
%%  &=& 1-Pr(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) ) \nonumber \\
%%  &=& 1-Pr(|t_{obs}| \le F_{|t|}^{-1}(1-p)) \nonumber \\
%%  &=&\left\{ \begin{array}{ll}
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p))=p &\mbox{ ; under $H_0$} \\
%%          1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%%       \end{array} \right.     
%%\end{eqnarray}
%%
%%Thus the density of $p_D$ is Uniform(0,1)  under $H_0$. As noted by \cite{Ruppert:2007}, under $H_1$ the density $$f_{p_D}(p_D; \delta)= \frac{f_{|t|}(F_{|t|}^{-1}(1-p_D);\delta)}{f_{|t|}(F_{|t|}^{-1}(1-p_D))}$$ derived from equation \eqref{dist_p}  is a right skewed distribution.  
%In a lineup plot we simulate $m-1$ residual data sets from null model where each of these $m-1$ null data sets produces a corresponding $p$-value $p_{0,i}$ and $p_{0,i} \sim \text{Uniform}(0,1)$ for $i = 1, ..., m-1$. Suppose  $p_0=\min_i(p_{0,i})$. Thus $p_0 \sim \text{Beta}(1,m-1)$.

@@

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% equation \eqref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \eqref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected_k.pdf}}
       \caption{Comparison of the expected power of a visual test of size $m=20$ for different $K$ (number of observers) with the power of the conventional test, for $n =100$ and $\sigma = 12$.  
       }
       \label{fig:power_expected}
\end{figure}

% \pagebreak

%\subsection{Some Power Considerations}

%\subsection{Expected number of choices}
%Since $X$ has a binomial distribution, we can have a look at the number of plots among the null plots that we would expect to be picked over the data plot:
%\[
%E[X] = (m-1)(1-E[p_D]).
%\]
%With an increase of the lineup size $m$ the number of null plots with a potentially stronger signal than the data plot grows linearly.
%This should allow us to infer some of the signal strength $p_D$ as
%\begin{equation} \label{plot_signal}
%E[p_D] = 1 - E[X]/(m-1),
%\end{equation}
%i.e. by averaging over the number of plots with  a stronger signal than the data plot we can evaluate signal strength of the plot even in the case where the data plot is not being picked by the observer.
%

@@

Participants for all the experiments were recruited through Amazon's online web service, Mechanical Turk  \citep{turk}. A summary of the data obtained for all three experiments are shown in Table \ref{tbl:summary}. Participants were asked to select the plot they think best matched the question given, provide a reason for their choice, and say how confident they are in their choice. Gender, age, education and geographic location of each participant are also collected. %The data was collected through the web. 
For each of the experiments one of the lineups was used as a test plot (easy plot) which everyone should get correct, so that a measure of the quality of the subjects effort could be made. Note that no participant was shown the same lineup twice.
%\green{Something probably needs to be said about subjects who do multiple tests, what checks are done.} \blue{We do have some data where we notice multiple responses for a lineup by a subject. This happens when subject accidentally use browser's ``back'' button and same data is saved twice. We made sure in the data cleaning process that we do not include more than one response on a lineup by a subject. For this we only keep the very first response.}

@@

%{\bf Steps to obtain lineup}
%\begin{enumerate}
%	\item For a given parameter setting obtain actual data set
%	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
%	\item Generate actual data plot using the procedure as described in table \ref{tbl:stat_multiple}
%	\item Generate 19 null (residual) data sets from $N(0, {\hat{\sigma}}^2$) and obtain 19 null plots
%	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
%\end{enumerate}

The number of evaluations required for each lineup to provide reasonable estimates of the proportion correct ($\hat p$) is determined by the variance of the number of correct evaluations. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) we consider $\gamma =p$.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

For each combination of sample size and standard deviation, each participant is given five randomly selected lineups, one of each difficulty level. Another set of four lineups is chosen from a second tier of selected combinations of sample size and standard deviation, with difficulty levels 0 to 3.  A last lineup was randomly selected from a set of  lineups with difficulty level 0. The order in which the lineups are shown to participants is randomized. %Each subject saw a substantial number of difficult lineups.

@@

%{\bf Steps to obtain lineup}
%\begin{enumerate}
%	\item For a given parameter setting obtain actual data set
%	\item Fit the null model to the actual data and obtain residuals and other parameter estimates ($\hat{\beta}$ and $\hat{\sigma}$)
%	\item Generate actual data plot (a scatter plot of Residual vs $X_1$)
%	\item Generate 19 null residual data sets from $N(0, {\hat{\sigma}}^2$) and obtain 19 null plots
%	\item Finally randomly place 19 null plots and one actual plot in a layout (4X5) to get a lineup 
%\end{enumerate}

@@

%\begin{figure}[hbt]
%\begin{figurehere}
%   \centering
%       \scalebox{0.6}{\includegraphics{contaminated_data.pdf}}
%       \caption{Scatterplot of a simulated contaminated data set generated such that a simple linear model produce a slope almost negligible, ie, if we fit model (Equation \ref{multi}) with $p=1$  to this data, the resulting estimate of $\beta_1$ would not be statistically significant. The conventional t-test $p$-value for the slope is 0.67 for this data set but visually we can see a convincing evidence of positive slope. 
%}
%       \label{fig:cont_dat}
%\end{figure}

@@

%\green{Fill in details of lineups, probably need a figure with an example lineup, and how lineups were assigned to subjects. Show a sample lineup, probably the most difficult, instead of the single plot.}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed.pdf}}
%       \caption{Observed power of visual test from equation \eqref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
%       \label{fig:power_observed}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.7}{\includegraphics{power_loess_exp1.pdf}}
%       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 1.}
%       \label{fig:power_loess1}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.7}{\includegraphics{power_loess_exp2.pdf}}
%       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of UMP test for sample size $n= 100,300$ and $\sigma = 12,5$ as per experiment 2.}
%       \label{fig:power_loess2}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.7}{\includegraphics{power_diff_exp.pdf}}
%       \caption{Difference of Observed power from UMP test power obtained in three experiments for sample size $n= 100,300$ and $\sigma = 12,5$ and values of slope parameter $\beta$ as shown in Table \ref{tbl:experiment_params}.}
%       \label{fig:power_diff_exp}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.6}{\includegraphics{power_observed_exp3.pdf}}
%       \caption{Observed power of visual test from equation \eqref{eqn:power_estimate} with pointwise 95\% confidence limits and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. }
%       \label{fig:power_observed_exp3}
%\end{figure*}

@@

%{\bf Screening scenarios:}
Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage in the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we use one of the very easy lineups that everybody was exposed to as a {\it reference lineup} and  take action based on a subject's answer to this reference:
if the subject failed to identify the actual data plot on the reference lineup, we remove all of this subject's data from the analysis. If the answer on the reference lineup is correct, we remove the answer for this lineup from the analysis, but keep all of the remaining answers. Table~\ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  A summary of all the data along with cleaned data are added to the supplementary material.

@@

 One assumption made in order to evaluate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value.  We examine  the data collected from the first two experiments, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figure \ref{fig:P-val_log2} gives an overview of all selections in all lineups of experiments 1 and 2. Each panel of the figure corresponds to a single lineup. Each `pin' -- a short line topped by a dot -- corresponds to one plot in the lineup. The $x$-location of the pin shows the plot's $p$-value on a log scale, its height  is given by the number of observer choosing this plot.
%Figure \ref{fig:P-val_log2} shows the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. 
Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

>>>>

@@

@@

Several new developments in graphics research have been achieved in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. The research by \cite{heer:2010} used subjects recruited from Amazon's Mechanical Turk \citep{turk} for their studies. This body of work provides a contemporary framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual~$i$ picks out the actual data plot. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot.

If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p_i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition (see \citet{butler93} for details). Ways to estimate $p_i$ will be  discussed in the following sections.

\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test, $V_{\theta}$, is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0 \mid \theta) 
    \end{equation*}
\end{dfn}

@@

The above lemma allows two immediate conclusions for the use of lineups.
The probability that the observer correctly identifies the data plot is closely connected to the size of the lineup $m$, since the right hand side of the above equation decreases for larger $m$, the probability of correctly identifying the actual data plot decreases with $m$. Further we see, that the rate of this decrease depends strongly on the distribution of $p_D$ -- if the density of $p_D$ is very right skewed, the  expectation term on right hand side will be large and less affected by an increase in $m$. This can also be seen in Figure \ref{fig:pval_power}, which illustrates lemma \ref{lemma}.   Figure \ref{fig:pval_power} shows the probability of picking the actual data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. It can also be seen that for a $p$-value, $p_D$, of about 0.15 for the data plot, the signal in the plot is so weak, that it cannot be distinguished from null plots in a lineup of size $m=20$.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. }
   \label{fig:pval_power}
\end{figure}

@@

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected_k.pdf}}
       \caption{Comparison of the expected power of a visual test of size $m=20$ for different $K$ (number of observers) with the power of the conventional test, for $n =100$ and $\sigma = 12$.  
       }
       \label{fig:power_expected}
\end{figure}

@@

Participants for all the experiments were recruited through Amazon's online web service, Mechanical Turk  \citep{turk}. A summary of the data obtained for all three experiments are shown in Table \ref{tbl:summary}. Participants were asked to select the plot they think best matched the question given, provide a reason for their choice, and say how confident they are in their choice. Gender, age, education and geographic location of each participant are also collected. %The data was collected through the web. 
For each of the experiments one of the lineups was used as a test plot (easy plot) which everyone should get correct, so that a measure of the quality of the subjects effort could be made. Note that no participant was shown the same lineup twice.

%\green{Something probably needs to be said about subjects who do multiple tests, what checks are done.} \blue{We do have some data where we notice multiple responses for a lineup by a subject. This happens when subject accidentally use browser's ``back'' button and same data is saved twice. We made sure in the data cleaning process that we do not include more than one response on a lineup by a subject. For this we only keep the very first response.}

@@

The number of evaluations required for each lineup to provide reasonable estimates of the proportion correct ($\hat p$) is determined by the variance of the number of correct evaluations. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) we consider $\gamma =p$.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

@@

For each combination of sample size and standard deviation, each participant is given five randomly selected lineups, one of each difficulty level. Another set of four lineups is chosen from a second tier of selected combinations of sample size and standard deviation, with difficulty levels 0 to 3.  A last lineup was randomly selected from a set of  lineups with difficulty level 0. The order in which the lineups are shown to participants is randomized.

@@

@@

@@

@@

Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage in the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we use one of the very easy lineups that everybody was exposed to as a {\it reference lineup} and  take action based on a subject's answer to this reference:
if the subject failed to identify the actual data plot on the reference lineup, we remove all of this subject's data from the analysis. If the answer on the reference lineup is correct, we remove the answer for this lineup from the analysis, but keep all of the remaining answers. Table~\ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  A summary of all the data along with cleaned data are added to the supplementary material.

@@

 One assumption made in order to evaluate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value.  We examine  the data collected from the first two experiments, to see if this assumption is, indeed, reasonable.

Figure \ref{fig:P-val_log2} gives an overview of all selections in all lineups of experiments 1 and 2. Each panel of the figure corresponds to a single lineup. Each `pin' -- a short line topped by a dot -- corresponds to one plot in the lineup. The $x$-location of the pin shows the plot's $p$-value on a log scale, its height  is given by the number of observer choosing this plot.
Columns are ordered according to effect size as defined in section 6.2; rows show replicates for the same combination of parameters.

Red indicates the plot with the lowest $p$-value in the lineup. Blue indicates the plot of the actual data when it is different from that with the lowest $p$-value. In both experiments  people tended to select the plot with the lowest $p$-value. The results are clearer for experiment 2, that used a continuous covariate. But even when subjects did not pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.
